---
title: "Longitudinal treatments and time-dependent confounding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{longitudinal-trt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ipeval)
library(survival)
```

# Model development

First, simulate model development data. We use scenario 1 (Cox model) as described in Keogh and Van Geloven (2024). Our validation data will be simulated later in the same way as the development data.

```{r}
n <- 3000
n_visits <- 5

gamma_0 <- -1
gamma_L <- 0.5

alpha_0 <- -2
alpha_A <- -0.5
alpha_L <- 0.5
alpha_U <- 0.5

simulate_longitudinal <- function(n, fix_trt = NULL) {
  U <- rnorm(n, 0, 0.1)

  A <- matrix(nrow = n, ncol = n_visits)
  L <- matrix(nrow = n, ncol = n_visits)

  time <- rep(NA, n)
  status <- rep(NA, n)

  simulate_A <- function(i, L, fix_trt) {
    if (is.null(fix_trt)) {
      return(rbinom(n, 1, plogis(gamma_0 + gamma_L * L[, i])))
    } else {
      return(rep(fix_trt, n))
    }
  }

  L[, 1] <- rnorm(n, U, 1)
  A[, 1] <- simulate_A(1, L, fix_trt)

  for (i in 2:n_visits) {
    L[, i] <- rnorm(n, 0.8 * L[, i - 1] - A[, i - 1] + 0.1 * (i-i) + U)
    A[, i] <- ifelse(A[, i - 1] == 1, rep(1, n), simulate_A(i, L, fix_trt))
  }

  for (i in 1:n_visits){
    new.t <- ipeval:::simulate_time_to_event(
      n = n,
      constant_baseline_haz = 1,
      LP = alpha_0 + alpha_A * A[, i] + alpha_L * L[, i] + alpha_U * U
    )
    time <- ifelse(is.na(time) & new.t < 1, i - 1 + new.t,time)
  }
  status <- ifelse(is.na(time), 0, 1)
  time <- ifelse(is.na(time), 5, time)

  colnames(A) <- paste0("A", 0:(n_visits - 1))
  colnames(L) <- paste0("L", 0:(n_visits - 1))

  data.frame(id = 1:n, time, status, A, L, U)
}

df_dev <- simulate_longitudinal(n)

# prepare into long format for fitting Cox model

df_dev_long <- wide_to_long(
  df_dev,
  baseline_variables = c("id", "time", "status", "L0", "U"),
  wide_variables = list(A = c("A0", "A1", "A2", "A3", "A4"),
                        L = c("L0", "L1", "L2", "L3", "L4")),
  visit_times = c(0,1,2,3,4),
  outcome_times = df_dev$time
)

df_dev_long$L0 <- ave(df_dev_long$L, df_dev_long$id, FUN = function(x) x[[1]])
df_dev_long$time_start <- df_dev_long$visit_time
df_dev_long$time_end <- pmin(df_dev_long$visit_time + 1, df_dev_long$time)
df_dev_long$status <- ifelse(
  df_dev_long$time_end == df_dev_long$time, 
  df_dev_long$status, 
  0
)
df_dev_long$time <- NULL

df_dev_long <- add_lag_terms(df_dev_long, "A", 1:4)

head(df_dev)

head(df_dev_long)
```

Now to fit an interventional prediction model capable of estimating risk under always treated strategy and risk under never treated strategy.

```{r}
# TODO: this code needs to be refactored & annotated
ipt_visits <- ipeval:::ipt_weights(df_dev_long, A ~ L * A_lag_1)
ipeval:::print_model(ipt_visits$model)

df_dev_long$iptw_visits <- ipt_visits$weights
df_dev_long$iptw_cumprod <- unlist(tapply(df_dev_long$iptw_visits, df_dev_long$id, cumprod))

cox_msm <- coxph(
  Surv(time_start, time_end, status) ~ A + A_lag_1 + A_lag_2 + A_lag_3 + A_lag_4 + L0,
  data = df_dev_long,
  weights = iptw_cumprod
)

cumhaz <- basehaz(cox_msm, centered = FALSE)$hazard
event.times <- basehaz(cox_msm, centered = FALSE)$time
cumhaz.fun <- stepfun(event.times, c(0, cumhaz))

# hazard rate term
hrt <- function(variable, value) {
  exp(coef(cox_msm)[variable]*value)
}

# # risk under never treated
risk0 <- function(t, L0) {
  1 - exp(-cumhaz.fun(t)*hrt("L0", L0))
}

# # risk under always treated
risk1 <- function(t, L0) {
  1 - exp(-(
    cumhaz.fun(min(t, 1))*
      hrt("L0", L0)*hrt("A", 1) +
      
      (t >= 1) *
      (cumhaz.fun(min(t, 2)) - cumhaz.fun(1))*
      hrt("L0", L0)*hrt("A", 1)*hrt("A_lag1", 1) +
      
      (t >= 2) *
      (cumhaz.fun(min(t, 3)) - cumhaz.fun(2))*
      hrt("L0", L0)*hrt("A", 1)*hrt("A_lag1", 1)*hrt("A_lag2", 1) +
      
      (t >= 3) *
      (cumhaz.fun(min(t, 4)) - cumhaz.fun(3))*
      hrt("L0", L0)*hrt("A", 1)*hrt("A_lag1", 1)*hrt("A_lag2", 1)*hrt("A_lag3", 1) +
      
      (t >= 4) *
      (cumhaz.fun(min(t, 5)) - cumhaz.fun(4))*
      hrt("L0", L0)*hrt("A", 1)*hrt("A_lag1", 1)*hrt("A_lag2", 1)*
      hrt("A_lag3", 1)*hrt("A_lag4", 1)
  ))
}
```

We can use ```risk0(t, L0)``` to estimate the risk under no treatment at time $t$ for a patient with baseline value $L0$. ```risk1(t, L0)``` estimates risk under treatment.

# Validation

This will show the workflow for validating such models when time dependent confounding is an issue. We first simulate a validation data set. We expect the user to start with a dataframe in this (wide) structure. Inputs required for the ip_score_long are:

- a data.frame with the surv(time, status) outcome and id columns
- a data.frame in long format, which will contain the treatment and confounders at each time visit  

The package provides functions to easily generate these data.frames, when starting with data in this wide structure: 

```{r}
df_val <- simulate_longitudinal(n = 3000)
head(df_val)
```

We naturally expect the user to have estimated probabilities corresponding to the intervention(s) of interest.
```{r}
predicted0 <- risk0(5, df_val$L0)
predicted1 <- risk1(5, df_val$L0)
```


First, create the data.frame with the surv status:
```{r}
df_outcomes <- df_val[, c("id", "time", "status")]
head(df_outcomes)
```

Second, create the data.frame with the time dependent variables. The package provides several convenience functions to help with this. Our treatment model depends on the value of L of the same visit, and previous treatment value.

```{r}
df_val_long <- wide_to_long(
  df_val, 
  baseline_variables = c("id", "L0"), 
  wide_variables = list("A" = paste0("A", 0:4), "L" = paste0("L", 0:4)), 
  visit_times = 0:4,
  outcome_times = df_val$time
)

df_val_long <- add_lag_terms(df_val_long, "A")
head(df_val_long)
```

We can then call the validation function:

```{r}
ip_score_long(
  probabilities = predicted0, 
  data_outcome = df_outcomes, 
  data_long = df_val_long, 
  time_horizon = 5, 
  treatment_formula = A ~ A_lag_1 * L,
  treatment_of_interest = rep(0,5),
)


```
